---
title: "EEG Biomarker Dashboard: MCI vs Alzheimers Disease"
author: "Youwei Hu"
date: "2026-03-06"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
    source_code: embed
    css: dashboard_style.css
    navbar:
      - { title: "GitHub", href: "https://github.com/hww228/data555_Dashboard", align: right, icon: "fa-github" }
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(readxl)
library(dplyr)
library(tidyr)
library(janitor)
library(ggplot2)
library(plotly)
library(forcats)
# -----------------------------
# Load and Clean Data
# -----------------------------
patient_result <- read_excel("data/patient_data.xlsx")

orig_names <- colnames(patient_result)

patient_result_clean <- patient_result %>%
  janitor::clean_names() %>%
  mutate(
    diagnosis = as.factor(diagnosis),
    gender    = as.factor(gender)
  )

# map from cleaned -> original display
name_map <- tibble(
  clean = janitor::make_clean_names(orig_names),
  display = orig_names
)

# -----------------------------
# Define feature families
# -----------------------------

band_power_abs <- c("delta_abs","theta_abs","alpha_abs","beta_abs")
band_power_rel <- c("delta_rel","theta_rel","alpha_rel","beta_rel")

ratios <- c(
  "slow_delta_theta",
  "fast_alpha_beta",
  "delta_alpha",
  "theta_alpha_delta",
  "delta_alpha_beta",
  "delta_theta_alpha_beta"
)

wavelet_coherence <- c("wc_global","wc_delta","wc_theta","wc_alpha","wc_beta")

wavelet_energy <- c(
  "global_energy",
  "rel_delta_ene","rel_theta_ene","rel_alpha_ene","rel_beta_ene"
)

complexity <- c(
  "complexity_total",
  "c_delta","c_theta","c_alpha","c_beta"
)

feature_family <- tibble(
  feature = c(band_power_abs, band_power_rel, ratios, wavelet_coherence, wavelet_energy, complexity),
  family = c(
    rep("Spectrum Absolute Power", length(band_power_abs)),
    rep("Spectrum Relative Power", length(band_power_rel)),
    rep("Power Ratios", length(ratios)),
    rep("Connectivity", length(wavelet_coherence)),
    rep("Wavelet Energy", length(wavelet_energy)),
    rep("Complexity", length(complexity))
  )
)


# -----------------------------
# Labels for cleaner display
# -----------------------------
simple_labels <- c(
  "delta_abs" = "log10 Absolute Delta Power (uV2/Hz)",
  "theta_abs" = "log10 Absolute Theta Power (uV2/Hz)",
  "alpha_abs" = "log10 Absolute Alpha Power (uV2/Hz)",
  "beta_abs"  = "log10 Absolute Beta Power (uV2/Hz)",

  "delta_rel" = "Relative Delta Power",
  "theta_rel" = "Relative Theta Power",
  "alpha_rel" = "Relative Alpha Power",
  "beta_rel"  = "Relative Beta Power",

  "slow_delta_theta" = "log10 Slow Band Power (Delta + Theta)",
  "fast_alpha_beta"  = "log10 Fast Band Power (Alpha + Beta)",

  "delta_alpha" = "Delta / Alpha Ratio",
  "theta_alpha_delta" = "(Theta + Alpha) / Delta Ratio",
  "delta_alpha_beta" = "Delta / (Alpha + Beta) Ratio",
  "delta_theta_alpha_beta" = "(Delta + Theta) / (Alpha + Beta) Ratio",

  "wc_global" = "Global Wavelet Coherence",
  "wc_delta" = "Delta Wavelet Coherence",
  "wc_theta" = "Theta Wavelet Coherence",
  "wc_alpha" = "Alpha Wavelet Coherence",
  "wc_beta" = "Beta Wavelet Coherence",

  "global_energy" = "Global Wavelet Energy",
  "rel_delta_ene" = "Relative Delta Wavelet Energy",
  "rel_theta_ene" = "Relative Theta Wavelet Energy",
  "rel_alpha_ene" = "Relative Alpha Wavelet Energy",
  "rel_beta_ene"  = "Relative Beta Wavelet Energy",

  "complexity_total"       = "Global EEG Complexity",
  "c_delta"                = "Delta Band Complexity",
  "c_theta"                = "Theta Band Complexity",
  "c_alpha"                = "Alpha Band Complexity",
  "c_beta"                 = "Beta Band Complexity"
)

log_features <- c(
  "delta_abs", "theta_abs", "alpha_abs", "beta_abs",
  "slow_delta_theta", "fast_alpha_beta"
)



# -----------------------------
# Build Helper Function
# -----------------------------

to_display <- function(clean_names_vec) {
  out <- name_map$display[match(clean_names_vec, name_map$clean)]
  ifelse(is.na(out), clean_names_vec, out)
}

to_clean <- function(display_name) {
  out <- name_map$clean[match(display_name, name_map$display)]
  ifelse(is.na(out), display_name, out)
}

selected_features_by_family <- function(families) {
  feature_family %>%
    filter(family %in% families) %>%
    pull(feature)
}

calc_p_value <- function(dat, feature_col) {
  tmp <- dat %>%
    select(diagnosis, all_of(feature_col)) %>%
    drop_na()

  if (nrow(tmp) == 0) return(NA_real_)
  if (length(unique(tmp$diagnosis)) != 2) return(NA_real_)

  tryCatch(
    wilcox.test(tmp[[feature_col]] ~ tmp$diagnosis)$p.value,
    error = function(e) NA_real_
  )
}

make_corr <- function(dat, features, method = "pearson") {
  X <- dat %>%
    select(all_of(features)) %>%
    mutate(across(everything(), as.numeric))

  # Keep columns with at least 3 non-missing values and non-zero variance
  keep_cols <- names(X)[sapply(X, function(x) sum(!is.na(x)) >= 3 && sd(x, na.rm = TRUE) > 0)]
  X <- X %>% select(all_of(keep_cols))

  if (ncol(X) < 2) return(NULL)

  cor(X, use = "pairwise.complete.obs", method = method)
}

label_lookup <- function(x) {
  ifelse(x %in% names(simple_labels), unname(simple_labels[x]), x)
}
```

# Biomarker Comparison

## Selection {.sidebar data-width="300"}

```{r boxplot-selection-sidebar}

checkboxGroupInput(
  "family_box",
  "Feature family",
  choices = unique(feature_family$family),
  selected = "Spectrum Absolute Power"
)


renderUI({
  req(input$family_box)
  available_features <- selected_features_by_family(input$family_box)

  selectInput(
    "feature_box",
    "Select biomarker",
    choices = stats::setNames(available_features, label_lookup(available_features)),
    selected = available_features[1]
  )
})


selectInput(
  "gender_box",
  "Gender filter",
  choices = c("All", levels(patient_result_clean$gender)),
  selected = "All"
)

sliderInput(
  "age_range_box",
  "Age range",
  min = floor(min(patient_result_clean$age, na.rm = TRUE)),
  max = ceiling(max(patient_result_clean$age, na.rm = TRUE)),
  value = c(floor(min(patient_result_clean$age, na.rm = TRUE)), ceiling(max(patient_result_clean$age, na.rm = TRUE))),
  step = 1
)

checkboxInput("show_points", "Show individual points", FALSE)

```

## Plot {data-width="700"}

### Distribution of Selected Biomarker {data-height="800"}

```{r render-boxplot}

renderPlotly({
  req(input$feature_box)

  d <- patient_result_clean %>%
    filter(age >= input$age_range_box[1], age <= input$age_range_box[2])

  if (input$gender_box != "All") {
    d <- d %>% filter(gender == input$gender_box)
  }

  feature_col <- input$feature_box
  feature_label <- if (feature_col %in% names(simple_labels)) {
    unname(simple_labels[feature_col])
  } else {
    feature_col
  }

  should_log <- input$feature_box %in% log_features

  plot_dat <- d %>%
    select(patient_id, diagnosis, gender, age, all_of(feature_col)) %>%
    mutate(
      diagnosis = factor(diagnosis, levels = c("MCI", "AD")),
      raw_value = .data[[feature_col]],
      value_plot = if(should_log) log10(raw_value + 1) else as.numeric(raw_value)
    ) 
  
  validate(
    need(nrow(plot_dat) > 0, "No data available after filtering.")
  )
    
  p_val <- calc_p_value(d, feature_col)
  y_lab <- if (should_log) {
    "log transferred biomarker value"
  } else {
    "biomarker value"
  }
  
  p <- ggplot(
    plot_dat,
    aes(
      x = diagnosis,
      y = value_plot,
      fill = diagnosis,
      text = paste0(
        "Patient ID: ", patient_id, "<br>",
        "Diagnosis: ", diagnosis, "<br>",
        "Gender: ", gender, "<br>",
        "Age: ", age, "<br>",
        "Raw value: ", round(raw_value, 3), "<br>",
        "Displayed value: ", round(value_plot, 3)
      )
    )
  ) +
    geom_boxplot(
      width = 0.6,
      outlier.alpha = 0.6,
      color = "black"
    ) +
    {
      if (isTRUE(input$show_points)) {
        geom_jitter(
          width = 0.10,
          size = 2.1,
          alpha = 0.70,
          color = "gray25"
        )
      }
    } +
    scale_fill_manual(values = c("MCI" = "#2C7BB6", "AD"  = "#D7191C"),
                      labels = c("MCI" = "Mild Cognitive Impairment (MCI)",
                                 "AD"  = "Alzheimer's Disease (AD)")) +
    labs(
      x = NULL,
      y = y_lab,
      fill = NULL
    ) +
    theme_bw(base_size = 14) +
    theme(
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = 14),
      axis.title.x  = element_text(size = 14, margin = margin(t = 8)),
      axis.title.y  = element_text(size = 13),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11)
    )
  
  p_label <- if (is.na(p_val)) {
    "p-value: N/A"
  } else if (p_val < 0.001) {
    "Wilcoxon p < 0.001"
  } else {
    paste0("Wilcoxon p = ", signif(p_val, 3))
  }

  ggplotly(p, tooltip = "text", height = 550) %>%
  layout(
    title = list(
      text = paste0(
        "Distribution of ", feature_label, " by Diagnosis Group</b>",
        "<br><span style='font-size:14px;color:#555;'>", p_label, "</span>"),
      x = 0.5,
      font = list(size = 16)
    ),
    margin = list(t = 120)
  )
  
  
})
```

### Key Takeaways from Boxplot {data-height="200"}

These boxplots compare the distribution of a single EEG biomarker between patients with **Mild Cognitive Impairment (MCI)** and **Alzheimer's Disease (AD)**, with optional filters for age and gender. The box shows the interquartile range (IQR), and the line marks the median. Hovering over any point or box reveals the patient ID, diagnosis, gender, age, and the precise biomarker value — making it easy to inspect individual observations and outliers. The Wilcoxon rank-sum p-value beneath the title tests whether the two group distributions differ significantly.

For absolute-power and slow/fast-band features, values are displayed on a log₁₀ scale to reduce skew; the original raw value is always available in the tooltip.

# Correlation Explorer

## Selection {.sidebar data-width="300"}

```{r plot-two-correlation-heatmap}
checkboxGroupInput(
  "family_corr",
  "Feature families for heatmap",
  choices = unique(feature_family$family),
  selected = unique(feature_family$family)
)

selectInput(
  "diag_corr",
  "Diagnosis subset",
  choices = c("All", levels(patient_result_clean$diagnosis)),
  selected = "All"
)

selectInput(
  "gender_corr",
  "Gender subset",
  choices = c("All", levels(patient_result_clean$gender)),
  selected = "All"
)

sliderInput(
  "age_range_corr",
  "Age Range (years)",
  min = floor(min(patient_result_clean$age, na.rm = TRUE)),
  max = ceiling(max(patient_result_clean$age, na.rm = TRUE)),
  value = c(
    floor(min(patient_result_clean$age, na.rm = TRUE)),
    ceiling(max(patient_result_clean$age, na.rm = TRUE))
    ),
  step = 1
)

selectInput(
  "corr_method",
  "Correlation method",
  choices = c("pearson", "spearman"),
  selected = "pearson"
)

sliderInput(
  "corr_thr",
  "Show |correlation| ≥",
  min = 0,
  max = 0.95,
  value = 0.30,
  step = 0.05
)
```

## Plot {data-width="700"}

### Correlation Heatmap {data-height="800"}

```{r correlation-heatmap}
corr_data <- reactive({
  d <- patient_result_clean %>%
    filter(age >= input$age_range_corr[1], age <= input$age_range_corr[2])

  if (input$diag_corr != "All") {
    d <- d %>% filter(diagnosis == input$diag_corr)
  } else {
    d <- d %>% filter(!is.na(diagnosis))
  }

  if (input$gender_corr != "All") {
    d <- d %>% filter(gender == input$gender_corr)
  }

  d
})

renderPlotly({
  req(input$family_corr)
  req(length(input$family_corr) > 0)
  req(input$corr_method)
  
  feats <- selected_features_by_family(input$family_corr)
  
  validate(
    need(length(feats) > 0, "No features available for the selected family.")
  )
  
  C <- make_corr(corr_data(), feats, method = input$corr_method)
  validate(
    need(!is.null(C), "Not enough valid features/data to compute correlations."),
    need(nrow(C) >= 2 && ncol(C) >= 2, "Need at least two valid features.")
  )
  
  
  C_show <- C
  C_show[abs(C_show) < input$corr_thr] <- NA
  x_labels <- colnames(C_show)
  y_labels <- rownames(C_show)

  x_labels_disp <- ifelse(
    x_labels %in% names(simple_labels),
    unname(simple_labels[x_labels]),
    x_labels
  )

  y_labels_disp <- ifelse(
    y_labels %in% names(simple_labels),
    unname(simple_labels[y_labels]),
    y_labels
  )
  
  plot_ly(
    x = x_labels_disp,
    y = y_labels_disp,
    z = C_show,
    type = "heatmap",
    source = "heatmap_click",
    hovertemplate = paste0(
      "<b>Biomarker X:</b> %{x}<br>",
      "<b>Biomarker Y:</b> %{y}<br>",
      "<b>Correlation (", input$corr_method, "):</b> %{z:.3f}",
      "<extra></extra>")
  ) %>%
    layout(
      title = paste0(
        "Correlation Heatmap (", input$corr_method,
        "), showing |r| ≥ ", input$corr_thr
      ),
      xaxis = list(
        title = list(text = "EEG Biomarker", font = list(size = 13)), 
        tickangle = 45),
      yaxis = list(title = "")
    )
})


```

### Key takeaways from correlation heatmap {data-height="200"}

This heatmap reveals clusters of EEG biomarkers that co-vary across patients. Strong positive correlations (blue) suggest redundant features that carry overlapping information; strong negative correlations (red) indicate features that move in opposite directions and may together capture complementary aspects of brain activity. Features with weak or no correlations are likely to contribute unique signal to any downstream classification or modeling task.

Use the filters to test whether the correlation structure changes across diagnosis group, gender, or age range — this can expose subgroup-specific patterns not visible in the full sample.


# About the Data

## Column

### Dataset Description {data-height="400"}

**Data Source:** Resting-state EEG recordings collected at the Goizueta Brain Health Institute at Emory University (data confidential; not publicly available).

**Sample Size:** 31 participants clinically diagnosed with either Alzheimer's Disease (AD) or Mild Cognitive Impairment (MCI) based on standard neurological assessments .

**Data Collection:** EEG signals were recorded under resting-state conditions — participants remained relaxed without performing any cognitive task. Signals were captured from 19 scalp electrodes (Fp1, Fp2, F3, F4, F7, F8, Fz, C3, C4, Cz, P3, P4, Pz, T3, T4, T5, T6, O1, O2) positioned according to the international 10–20 system, sampled at 256 Hz. Features extracted include spectral power, power ratios, wavelet coherence, wavelet energy, and complexity measures across the delta, theta, alpha, and beta frequency bands.

**Study Population:** Adults diagnosed with AD or MCI, evaluated through standard neurological assessment at the Goizueta Brain Health Institute, Emory University.


### Real-World Importance {data-height="200"}
Early and objective differentiation between Mild Cognitive Impairment and Alzheimer's Disease is critical for timely clinical intervention, yet current diagnosis relies heavily on subjective assessments. This dashboard enables clinicians and researchers to interactively explore which resting-state EEG biomarkers most reliably distinguish MCI from AD, supporting the development of accessible, non-invasive neurophysiological tools for earlier dementia detection.

*[GitHub Repository](https://github.com/hww228/data555_Dashboard) — includes source code and a full README.*

