---
title: "Data555 Dashboard"
author: "Youwei Hu"
date: "2026-03-06"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
    source_code: embed
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(readxl)
library(dplyr)
library(tidyr)
library(janitor)
library(ggplot2)
library(plotly)
library(forcats)
# -----------------------------
# Load and Clean Data
# -----------------------------
patient_result <- read_excel("data/patient_data.xlsx")

orig_names <- colnames(patient_result)

patient_result_clean <- patient_result %>%
  janitor::clean_names() %>%
  mutate(
    diagnosis = as.factor(diagnosis),
    gender    = as.factor(gender)
  )

# map from cleaned -> original display
name_map <- tibble(
  clean = janitor::make_clean_names(orig_names),
  display = orig_names
)

# -----------------------------
# Define feature families
# -----------------------------
# Adjust these lists if you want different grouping
band_power_abs <- c("delta_abs","theta_abs","alpha_abs","beta_abs")
band_power_rel <- c("delta_rel","theta_rel","alpha_rel","beta_rel")

ratios <- c(
  "slow_delta_theta",
  "fast_alpha_beta",
  "delta_alpha",
  "theta_alpha_delta",
  "delta_alpha_beta",
  "delta_theta_alpha_beta"
)

wavelet_coherence <- c("wc_global","wc_delta","wc_theta","wc_alpha","wc_beta")

wavelet_energy <- c(
  "global_energy",
  "rel_delta_ene","rel_theta_ene","rel_alpha_ene","rel_beta_ene"
)

complexity <- c(
  "complexity_total",
  "c_delta","c_theta","c_alpha","c_beta"
)

feature_family <- tibble(
  feature = c(band_power_abs, band_power_rel, ratios, wavelet_coherence, wavelet_energy, complexity),
  family = c(
    rep("Spectrum Absolute Power", length(band_power_abs)),
    rep("Spectrum Relative Power", length(band_power_rel)),
    rep("Power Ratios", length(ratios)),
    rep("Connectivity", length(wavelet_coherence)),
    rep("Wavelet Energy", length(wavelet_energy)),
    rep("Complexity", length(complexity))
  )
)


# -----------------------------
# Labels for cleaner display
# -----------------------------
simple_labels <- c(
  "delta_abs" = "log10 absolute delta",
  "theta_abs" = "log10 absolute theta",
  "alpha_abs" = "log10 absolute alpha",
  "beta_abs"  = "log10 absolute beta",

  "delta_rel" = "relative delta",
  "theta_rel" = "relative theta",
  "alpha_rel" = "relative alpha",
  "beta_rel"  = "relative beta",

  "slow_delta_theta" = "log10 slow (delta + theta)",
  "fast_alpha_beta"  = "log10 fast (alpha + beta)",

  "delta_alpha" = "delta/alpha ratio",
  "theta_alpha_delta" = "(theta + alpha)/delta",
  "delta_alpha_beta" = "delta/(alpha + beta)",
  "delta_theta_alpha_beta" = "(delta + theta)/(alpha + beta) ratio",

  "wc_global" = "global wavelet coherence",
  "wc_delta" = "delta wavelet coherence",
  "wc_theta" = "theta wavelet coherence",
  "wc_alpha" = "alpha wavelet coherence",
  "wc_beta" = "beta wavelet coherence",

  "global_energy" = "global wavelet energy",
  "rel_delta_ene" = "relative delta energy",
  "rel_theta_ene" = "relative theta energy",
  "rel_alpha_ene" = "relative alpha energy",
  "rel_beta_ene" = "relative beta energy",

  "complexity_total" = "global complexity",
  "c_delta" = "delta complexity",
  "c_theta" = "theta complexity",
  "c_alpha" = "alpha complexity",
  "c_beta" = "beta complexity"
)

log_features <- c(
  "delta_abs", "theta_abs", "alpha_abs", "beta_abs",
  "slow_delta_theta", "fast_alpha_beta"
)



# -----------------------------
# Build Helper Function
# -----------------------------

to_display <- function(clean_names_vec) {
  out <- name_map$display[match(clean_names_vec, name_map$clean)]
  ifelse(is.na(out), clean_names_vec, out)
}

to_clean <- function(display_name) {
  out <- name_map$clean[match(display_name, name_map$display)]
  ifelse(is.na(out), display_name, out)
}

selected_features_by_family <- function(families) {
  feature_family %>%
    filter(family %in% families) %>%
    pull(feature)
}

calc_p_value <- function(dat, feature_col) {
  tmp <- dat %>%
    select(diagnosis, all_of(feature_col)) %>%
    drop_na()

  if (nrow(tmp) == 0) return(NA_real_)
  if (length(unique(tmp$diagnosis)) != 2) return(NA_real_)

  tryCatch(
    wilcox.test(tmp[[feature_col]] ~ tmp$diagnosis)$p.value,
    error = function(e) NA_real_
  )
}

make_corr <- function(dat, features, method = "pearson") {
  X <- dat %>%
    select(all_of(features)) %>%
    mutate(across(everything(), as.numeric))

  # Keep columns with at least 3 non-missing values and non-zero variance
  keep_cols <- names(X)[sapply(X, function(x) sum(!is.na(x)) >= 3 && sd(x, na.rm = TRUE) > 0)]
  X <- X %>% select(all_of(keep_cols))

  if (ncol(X) < 2) return(NULL)

  cor(X, use = "pairwise.complete.obs", method = method)
}

label_lookup <- function(x) {
  ifelse(x %in% names(simple_labels), unname(simple_labels[x]), x)
}
```

# Biomarker Comparison

## Selection {.sidebar data-width="300"}

```{r boxplot-selection-sidebar}

checkboxGroupInput(
  "family_box",
  "Feature family",
  choices = unique(feature_family$family),
  selected = "Spectrum Absolute Power"
)


renderUI({
  req(input$family_box)
  available_features <- selected_features_by_family(input$family_box)

  selectInput(
    "feature_box",
    "Select biomarker",
    choices = stats::setNames(available_features, label_lookup(available_features)),
    selected = available_features[1]
  )
})


selectInput(
  "gender_box",
  "Gender filter",
  choices = c("All", levels(patient_result_clean$gender)),
  selected = "All"
)

sliderInput(
  "age_range_box",
  "Age range",
  min = floor(min(patient_result_clean$age, na.rm = TRUE)),
  max = ceiling(max(patient_result_clean$age, na.rm = TRUE)),
  value = c(floor(min(patient_result_clean$age, na.rm = TRUE)), ceiling(max(patient_result_clean$age, na.rm = TRUE))),
  step = 1
)

checkboxInput("show_points", "Show individual points", FALSE)

```

## Plot {data-width="700"}

### Distribution of Selected Biomarker {data-height="800"}

```{r render-boxplot}

renderPlotly({
  req(input$feature_box)

  d <- patient_result_clean %>%
    filter(age >= input$age_range_box[1], age <= input$age_range_box[2])

  if (input$gender_box != "All") {
    d <- d %>% filter(gender == input$gender_box)
  }

  feature_col <- input$feature_box
  feature_label <- if (feature_col %in% names(simple_labels)) {
    unname(simple_labels[feature_col])
  } else {
    feature_col
  }

  should_log <- input$feature_box %in% log_features

  plot_dat <- d %>%
    select(patient_id, diagnosis, gender, age, all_of(feature_col)) %>%
    mutate(
      diagnosis = factor(diagnosis, levels = c("MCI", "AD")),
      raw_value = .data[[feature_col]],
      value_plot = if(should_log) log10(raw_value + 1) else as.numeric(raw_value)
    ) 
  
  validate(
    need(nrow(plot_dat) > 0, "No data available after filtering.")
  )
    
  p_val <- calc_p_value(d, feature_col)
  y_lab <- if (should_log) {
    "log transferred biomarker value"
  } else {
    "biomarker value"
  }
  
  p <- ggplot(
    plot_dat,
    aes(
      x = diagnosis,
      y = value_plot,
      fill = diagnosis,
      text = paste0(
        "Patient ID: ", patient_id, "<br>",
        "Diagnosis: ", diagnosis, "<br>",
        "Gender: ", gender, "<br>",
        "Age: ", age, "<br>",
        "Raw value: ", round(raw_value, 4), "<br>",
        "Displayed value: ", round(value_plot, 4)
      )
    )
  ) +
    geom_boxplot(
      width = 0.6,
      outlier.alpha = 0.6,
      color = "black"
    ) +
    {
      if (isTRUE(input$show_points)) {
        geom_jitter(
          width = 0.10,
          size = 2.1,
          alpha = 0.70,
          color = "gray25"
        )
      }
    } +
    scale_fill_manual(values = c("MCI" = "#2C7BB6", "AD"  = "#D7191C")) +
    labs(
      x = NULL,
      y = y_lab,
      fill = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 15),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11)
    )

  
  ggplotly(p, tooltip = "text", height = 550) %>%
  layout(
    boxmode = "group",
    title = list(
      text = paste0(
        "Distribution of ", feature_label, " by Diagnosis",
        "<br><span style='font-size:14px;'>Wilcoxon p-value: ",
        ifelse(is.na(p_val), "NA", signif(p_val, 3)),
        "</sup>"
      ),
      x = 0.5
    ),
    margin = list(t = 120)
  )
  
  
})
```

### Key Takeaways from Boxplot {data-height="200"}

These boxplots compare the distribution of one EEG biomarker at a time between patients with mild cognitive impairment (Group MCI) and Alzheimer's disease (Group AD) after filtering by age and gender. The boxplot and overlaid patient-level points make it easy to see whether the AD group has a higher center, wider spread, or more extreme values than the MCI group, while the tooltip helps the viewer inspect individual observations and outliers in detail.

For absolute power and slow/fast power features, the displayed scale is `log10(value + 1)` so that highly skewed values remain readable. This makes group differences easier to compare visually while still preserving the original patient-level values in the tooltip.

# Correlation Explorer

## Selection {.sidebar data-width="300"}

```{r plot-two-correlation-heatmap}
checkboxGroupInput(
  "family_corr",
  "Feature families for heatmap",
  choices = unique(feature_family$family),
  selected = unique(feature_family$family)
)

selectInput(
  "diag_corr",
  "Diagnosis subset",
  choices = c("All", levels(patient_result_clean$diagnosis)),
  selected = "All"
)

selectInput(
  "gender_corr",
  "Gender subset",
  choices = c("All", levels(patient_result_clean$gender)),
  selected = "All"
)

sliderInput(
  "age_range_corr",
  "Age range for heatmap",
  min = floor(min(patient_result_clean$age, na.rm = TRUE)),
  max = ceiling(max(patient_result_clean$age, na.rm = TRUE)),
  value = c(floor(min(patient_result_clean$age, na.rm = TRUE)), ceiling(max(patient_result_clean$age, na.rm = TRUE))),
  step = 1
)

selectInput(
  "corr_method",
  "Correlation method",
  choices = c("pearson", "spearman"),
  selected = "pearson"
)

sliderInput(
  "corr_thr",
  "Show |correlation| ≥",
  min = 0,
  max = 0.95,
  value = 0.30,
  step = 0.05
)
```

## Plot {data-width="700"}

### Correlation Heatmap {data-height="800"}

```{r correlation-heatmap}
corr_data <- reactive({
  d <- patient_result_clean %>%
    filter(age >= input$age_range_corr[1], age <= input$age_range_corr[2])

  if (input$diag_corr != "All") {
    d <- d %>% filter(diagnosis == input$diag_corr)
  } else {
    d <- d %>% filter(!is.na(diagnosis))
  }

  if (input$gender_corr != "All") {
    d <- d %>% filter(gender == input$gender_corr)
  }

  d
})

renderPlotly({
  req(input$family_corr)
  req(length(input$family_corr) > 0)
  req(input$corr_method)
  
  feats <- selected_features_by_family(input$family_corr)
  
  validate(
    need(length(feats) > 0, "No features available for the selected family.")
  )
  
  C <- make_corr(corr_data(), feats, method = input$corr_method)

  validate(
    need(!is.null(C), "Not enough valid features/data to compute correlations."),
    need(nrow(C) >= 2 && ncol(C) >= 2, "Need at least two valid features to draw the heatmap.")
  )
  
  
  C_show <- C
  C_show[abs(C_show) < input$corr_thr] <- NA
  x_labels <- colnames(C_show)
  y_labels <- rownames(C_show)

  x_labels_disp <- ifelse(
    x_labels %in% names(simple_labels),
    unname(simple_labels[x_labels]),
    x_labels
  )

  y_labels_disp <- ifelse(
    y_labels %in% names(simple_labels),
    unname(simple_labels[y_labels]),
    y_labels
  )
  
  plot_ly(
    x = x_labels_disp,
    y = y_labels_disp,
    z = C_show,
    type = "heatmap",
    source = "heatmap_click",
    hovertemplate = paste0(
      "Feature X: %{x}<br>",
      "Feature Y: %{y}<br>",
      "Correlation: %{z:.3f}<extra></extra>"
    )
  ) %>%
    layout(
      title = paste0(
        "Correlation Heatmap (", input$corr_method,
        "), showing |r| ≥ ", input$corr_thr
      ),
      xaxis = list(title = "", tickangle = 45),
      yaxis = list(title = "")
    )
})


```

### Key takeaways from correlation heatmap {data-height="200"}

These correlation heatmap help identify clusters of biomarkers that move together and may therefore contain overlapping information. Strong positive or negative correlations suggest redundancy, while weaker correlations highlight features that may contribute unique information to downstream modeling or interpretation.

The filters let the viewer test whether the correlation structure changes across diagnosis group, gender, age range, feature family, and correlation method. This supports deeper exploration by allowing the user to move from a broad overview of the full biomarker set to a focused view of only the strongest relationships.
